---
title: "Random Forest"
author: "Harrison Cradduck, Nick Berry"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.width = 8, fig.height = 6)

if(require(pacman) == 0)
   {install.packages("pacman")}
pacman::p_load(devtools, caret, cluster, dplyr, fastDummies, leaps, pacman, tidyverse, skimr, fastDummies, GGally, DataExplorer, ggrepel, ggthemes, dslabs, lubridate, stringr, tidytext, purrr, forcats, rpart, ROCR, stringr, MLmetrics, doParallel, parallel)

if (!require(mlba)) {
  library(devtools)
  install_github("gedeck/mlba/mlba", force = TRUE)
}
pacman::p_load(mlba, tidyverse)
```

# Introduction

**Using the methods we discussed in class, develop the best Random Forest tree that you can.  Tune hyperparameters to improve model performance.**

# Loading in Data

```{r}
# read in training and holdout datsets edited specifically for our decision tree models
train <- readr:: read_rds('decision_tree_train.rds')

holdout <- readr::read_rds('decision_tree_holdout.rds')
```

```{r}
# set seed for ability to reproduce downsampling and modeling
set.seed(123)
```

# Downsampling

**For our Random Forest models we decided to move forward with just the downsampled data for the sake of computational complexity**

```{r}
# downsampling the majority class
train.ds <- caret::downSample(x = train %>% 
                    dplyr::select(-loan_default),
                  y = train$loan_default,
                  yname = "loan_default")
table(train.ds$loan_default)
```

# 1-Hot Dummy Encoding

## Training Data

```{r}
# creating the dummies on the training data 
dummies <- caret::dummyVars(~ . - loan_default, 
                            data = train.ds,
                            fullRank = FALSE)
# excluding the response variable so it remains in the same form in all our models
```

```{r}
# making the dataframe for the dummies and reintroducing the observed class
train.dummy <- data.frame(predict(dummies, newdata = train.ds))

train.dummy <- cbind(train.dummy, loan_default = train.ds$loan_default)
```

## Holdout Data

```{r}
# creating the dummies on the holdout data 
dummies <- caret::dummyVars(~ . - loan_default, 
                            data = holdout,
                            fullRank = FALSE)
# excluding the response variable so it remains in the same form in all our models
```

```{r}
# making the dataframe for the dummies and reintroducing the observed class
holdout.dummy <- data.frame(predict(dummies, newdata = holdout))

holdout.dummy <- cbind(holdout.dummy, loan_default = holdout$loan_default)
```

# Modeling

```{r}
#set up parallel processing for running the models
cores = parallel::detectCores()
cores

c1 <- parallel::makeCluster(8)
registerDoParallel(c1)
```

##Training a Random Forest on the Regular, Downsampled Data 


```{r}
# cross validating to select mtry parameter

# set up train control
cv_control <- trainControl(method = 'cv', number = 10, allowParallel = TRUE)

# define grid of mtry values to search over
mtry_grid <- expand.grid(mtry = seq(1, ncol(train.ds) - 1, by = 1))
```

**We used 10 fold cross validation to find the best mtry value from 1 to the number of variables in our downsampled data**

```{r}
# train the random forest model with cross validation
# rf_cv <- caret::train(loan_default ~ .,
#                       data = train.ds,
#                       method = 'rf',
#                       trControl = cv_control,
#                       tuneGrid = mtry_grid,
#                       ntree = 100,
#                       importance = TRUE,
#                       nodesize = 25)

# saveRDS(rf_cv, 'random_forest.rds')
```

```{r}
# read in the previously saved model
rf_cv <- readr::read_rds('random_forest.rds')
# print the model summary
rf_cv
```

**We fit our first random forest model on the down sampled training data. We used 100 trees with a minimum node size of 25. The mtry that was selected for the final model was 28.**

##Training a Random Forest on the 1-Hot Dummy Encoded, Downsampled Data

```{r}
# train the random forest model with cross validation on 1 hot encoded training downsampled data
# rf_cv_dmy <- caret::train(loan_default ~ .,
#                       data = train.dummy,
#                       method = 'rf',
#                       trControl = cv_control,
#                       tuneGrid = mtry_grid,
#                       ntree = 100,
#                       importance = TRUE,
#                       nodesize = 25)
#
#saveRDS(rf_cv_dmy, 'random_forest_dmy.rds')
```

```{r}
# read in the previously saved dummy encoded model
rf_cv_dmy <- read_rds('random_forest_dmy.rds')
# print the model summary
rf_cv_dmy
```

**We fit our second random forest model on 1-Hot dummy encoded training data. We also set the number of trees to 100 and minimum nodesize to 25 for this model. The mtry for this random forest model was 35**

```{r}
# stop parallel processsing
stopCluster(c1)
registerDoSEQ()
```

# Model Evaluation

```{r}
# creating a dataframe to house all the holdout predictions for these models and 
pred.holdout <- data.frame(loan_default = holdout$loan_default)
```

## Downsampled Model

```{r}
# save predicted probability for the holdout data
pred.holdout$rf_cv_prob <- predict(rf_cv,
                                   newdata = holdout,
                                   type = 'prob')[,'Yes']
```

```{r}
# computing ROC of curve
pred <- ROCR::prediction(pred.holdout$rf_cv_prob,
                         pred.holdout$loan_default)
perf <- ROCR::performance(pred, 'tpr', 'fpr')
```

```{r}
# plotting ROC curve
plot(perf, col = 'blue', lwd = 2, main = 'ROC Curve for Cross-Validated Training Model')
abline(a = 0, b = 1, col = 'red', lty = 2, lwd =2)
```

```{r}
# AUC of above curve
ROCR::performance(pred, 'auc')@y.values[[1]]
```

```{r}
# finding the optimal threshold

#finding precision and recall
prec <- ROCR::performance(pred, 'prec')
rec <- ROCR::performance(pred, 'rec')

# extracting precision and recall
precision <- prec@y.values[[1]]
recall <- rec@y.values[[1]]

#compute F1 
f1 = 2 * precision * recall / (precision + recall)
f1[is.nan(f1)] = 0

# combine F1 and cuttofs 
cutoffs = prec@x.values[[1]]
df_f1 = data.frame(f1, cutoffs)

# find the optimal F1 index, F1 score , and cutoff
opt_idx <- which.max(f1)
opt_f1 <- df_f1[opt_idx,]
opt_f1
```

```{r}
# using the max f1 score for the model to set the probability threshold save the predicted class based on that threshold. 
best_thresh <- 0.63
pred.holdout$rf_cv_class <- factor(
  ifelse(pred.holdout$rf_cv_prob >= best_thresh, 
         'Yes', 'No'))
```

```{r}
# print a confusion matrix for the random forest model
caret::confusionMatrix(pred.holdout$rf_cv_class,
                       pred.holdout$loan_default,
                       positive = 'Yes')
```

**The model trained on the regular downsampled data has an optimal predicted probability threshold of 0.63 based on maximizing the f1 score, which was 0.5922998. The other important metrics are as follow: AUC = 0.8504582, Accuracy = 0.819, Sensitivity = 0.6704, Specificity = 0.8552** 

```{r}
# variable importance
caret::varImp(rf_cv, type = 2)
```

```{r}
randomForest::varImpPlot(rf_cv$finalModel, type = 2)
```

**We evaluated the variable importance for the model and fount that years since last credit pull was by far the most important variable in the model with interest rate, debt settlement flag (yes), and dti being the next most important variables.**

## 1-Hot Dummy Encoded, Downsampled Model

```{r}
# save predicted probability for the dummy encoded holdout data
pred.holdout$rf_cv_dmy_prob <- predict(rf_cv_dmy,
                                   newdata = holdout.dummy,
                                   type = 'prob')[,'Yes']
```

```{r}
# computing ROC of curve
pred2 <- ROCR::prediction(pred.holdout$rf_cv_dmy_prob,
                         pred.holdout$loan_default)
perf2 <- ROCR::performance(pred, 'tpr', 'fpr')
```

```{r}
# plotting ROC curve
plot(perf2, col = 'blue', lwd = 2, main = 'ROC Curve for Cross-Validated Training Model')
abline(a = 0, b = 1, col = 'red', lty = 2, lwd =2)
```

```{r}
# AUC of above curve
ROCR::performance(pred2, 'auc')@y.values[[1]]
```

```{r}
# finding the optimal threshold

#finding precision and recall
prec <- ROCR::performance(pred2, 'prec')
rec <- ROCR::performance(pred2, 'rec')

# extracting precision and recall
precision <- prec@y.values[[1]]
recall <- rec@y.values[[1]]

#compute F1 
f1 = 2 * precision * recall / (precision + recall)
f1[is.nan(f1)] = 0

# combine F1 and cuttofs 
cutoffs = prec@x.values[[1]]
df_f1 = data.frame(f1, cutoffs)

# find the optimal F1 index, F1 score , and cutoff
opt_idx <- which.max(f1)
opt_f1 <- df_f1[opt_idx,]
opt_f1
```

```{r}
# using the max f1 score for the model to set the probability threshold save the predicted class based on that threshold. 
best_thresh <- 0.62
pred.holdout$rf_cv_dmy_class <- factor(
  ifelse(pred.holdout$rf_cv_dmy_prob >= best_thresh, 
         'Yes', 'No'))
```

```{r}
# print a confusion matrix for the random forest model
caret::confusionMatrix(pred.holdout$rf_cv_dmy_class,
                       pred.holdout$loan_default,
                       positive = 'Yes')
```
**The model trained on the regular downsampled data has an optimal predicted probability threshold of 0.62 based on maximizing the f1 score, which was 0.5935158. The other important metrics are as follow: AUC = 0.8508873, Accuracy = 0.8172, Sensitivity = 0.6805, Specificity = 0.8505**

```{r}
# variable importance
caret::varImp(rf_cv_dmy, type = 2)
```

```{r}
randomForest::varImpPlot(rf_cv_dmy$finalModel, type = 2)
```

**We evaluated the variable importance for the model and fount that years since last credit pull was by far the most important variable in the model with interest rate being the next most important before seeing another drastic drop off in importance.**

```{r}
# saving the holdout predictions to compare to future models
saveRDS(pred.holdout, 'holdout_predictions.rds')
```

# Model Comparison

```{r}
# comparing the important model metrics into one dataframe

# saving the observed classifications
truth <- pred.holdout$loan_default

# Get model names based on .prob columns
model_names <- names(pred.holdout) %>%
  stringr::str_subset("\\_prob$") %>%
  stringr::str_remove("\\_prob$")

# Function to compute metrics for each model
get_metrics <- function(model_name) {
  prob_col <- paste0(model_name, "_prob")
  class_col <- paste0(model_name, "_class")

  # Extract predictions
  prob <- pred.holdout[[prob_col]]
  pred <- factor(pred.holdout[[class_col]], levels = c("No", "Yes"))

  # Confusion matrix
  cm <- caret::confusionMatrix(pred, truth, positive = "Yes")

  # AUC
  roc_obj <- pROC::roc(truth, prob, quiet = TRUE)

  # Metrics
  data.frame(
    Model = model_name,
    Accuracy = cm$overall["Accuracy"],
    Sensitivity = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"],
    F1 = MLmetrics::F1_Score(
      y_pred = pred,
      y_true = truth,
      positive = "Yes"
    ),
    AUC = as.numeric(roc_obj$auc)
  )
}

# Apply to all models
results_df <- dplyr::bind_rows(lapply(model_names, get_metrics))

# Remove row names and display
rownames(results_df) <- NULL
print(results_df)
```

```{r}
# compare models by plotting ROC curves
roc_ds <- pROC::roc(holdout$loan_default, 
                    predict(rf_cv, 
                            holdout, 
                            type = "prob")[, "Yes"])
roc_ds_dmy <- pROC::roc(holdout.dummy$loan_default, 
                        predict(rf_cv_dmy, 
                                holdout.dummy, 
                                type = "prob")[, "Yes"])

plot(roc_ds, col = "blue", lwd = 2, main = "ROC Curves for Random Forest Models")
plot(roc_ds_dmy, add = TRUE, col = "red", lwd = 2)

legend("bottomright",
       legend = c(
         paste0("Downsampled AUC = ", 
                ROCR::performance(pred, 'auc')@y.values[[1]]),
         paste0("1Hot Dummy AUC = ", 
                ROCR::performance(pred2, 'auc')@y.values[[1]])
       ),
       col = c("blue", "red", "darkgreen"),
       lwd = 2,
       cex = 0.8)
```

**Both models perform similarly across all of the metrics that we are evaluating with the greatest difference being in sensitivity where the difference is 0.01. There is almost no visible deviation from the two ROC lines plotted above.**

# Final Model

**Since we are most concerned about maximizing the f1 score for out models in this project, the random forest model fit on the 1-hot dummy encoded performed just slighty better with and f1 of 0.5935158.**